### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
edits on the phone

In this paper, we will represent referring expressions (REs) as
formulas of description logic~\cite{baad:desc03}.  In order to make
this point, we will now first define the two description logics we
will be working with: \alc\ and \el.

\begin{definition}
  Formulas (or concepts) $\varphi$ of $\alc$ are generated by the
  following grammar:
$$
\varphi,\varphi' ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi'
\mid \exists R. \varphi
$$
where $p$ is in the set of propositional symbols \prop, $R$ is in the
set of relational symbols \rel. $\el$ is the negation-free fragment of $\alc$.

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$ where
$\Delta^\gM$ is a non-empy set and $\interp{\cdot}^\gM$ is an
interpretation function such that:
$$
\begin{array}{ccl}
\interp{p}^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
\interp{R}^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
\interp{\neg \varphi}^\gM & = & \Delta^\gM \backslash \interp{\varphi}^\gM\\
\interp{\varphi \sqcap \varphi'}^\gM & = & \interp{\varphi}^\gM \cap \interp{\varphi'}^\gM\\
\interp{\exists R.\varphi}^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}^\gM\\
& & \mbox{ and } i' \in \interp{\varphi}^\gM \}.\\
\end{array}
$$

Given a model $\gM$ and an individual $i$ in $\gM$, let
$\propm(i) = \{p \in \prop \mid i \in \interp{p}^\gM\}$.
\end{definition}

@@

Different description logics differ in the inventory of logical
connectives they permit.  The difference between \alc\ and \el\ is
that \alc\ permits negation and \el\ doesn't.  There are many other
description logics in the literature; some that we will get back to in
Section~\ref{sec:related} are $\mathcal{CL}$ (\el\ without existential
quantification, i.e., only conjunctions of atoms); $\mathcal{PL}$
(\alc\ without existential quantification, i.e., propositional logic);
and $\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic
negation).

A key notion in description logics that we will use here is
\emph{similarity} \todo{citation?}.  For any description logic $\gL$,
we will say that an individual $i$ is \emph{\gL-similar} to $i'$ in a
given model $\gM$ if for any formula $\varphi \in \gL$ such that $i
\in \interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
Intuitively, this means that there is no $\gL$-formula that would
distinguish $i$ from $i'$.  We say that the \emph{\gL-similarity set}
of some individual $i$ is the set of all individuals to which $i$ is
\gL-similar.

Notice that similarity is not necessarily a symmetrical relation: For
instance, $f_1$ is \el-similar to $f_2$ in
Fig.~\ref{fig:dale-haddock}b, but $f_2$ is not \el-similar to $f_1$
(it satisfies the formula $\exists \mathsf{in}.\mathsf{hat}$ and $f_1$
doesn't).  However, \alc-similarity is a symmetrical relation because
\alc\ contains negation; and indeed, $f_1$ is not \alc-similar to
$f_2$ either because it satisfied $\neg \exists
\mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
it is generally possible for some individual $a$ to be \el-similar but
not \alc-similar to some individual $b$, but not vice versa.

\todo{need to cite \cite{blac:moda01,kurt:expr99} in the appropriate
  places}

@@

\begin{definition}
Given a model $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
\subseteq \Delta^\gM \times \Delta^\gM$ such that 
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in \interp{R}^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in \interp{R}^\gM$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in \interp{R}^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in \interp{R}^\gM$ and
$(e_1',e_2') \in Z$.
\end{itemize}
%A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
%bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
\end{definition}

>>>>

In this paper, we will represent referring expressions as formulas of
description logic~\cite{baad:desc03}.  In order to make this point, we
will now define the two description logics we will be working with:
\alc\ and \el.

\emph{Formulas} (or \emph{concepts}) $\varphi$ of $\alc$ are generated
by the following grammar:
$$
\varphi,\varphi' ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi'
\mid \exists R. \varphi
$$
where $p$ is in the set of propositional symbols \prop, and $R$ is in
the set of relational symbols \rel. $\el$ is the negation-free
fragment of $\alc$.

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta,\interp{\cdot})$ where
$\Delta$ is a non-empy set and $\interp{\cdot}$ is an
interpretation function such that:
$$
\begin{array}{ccl}
\interp{p} & \subseteq & \Delta  \mbox{ for $p \in \prop$}\\
\interp{R} & \subseteq & \Delta \times \Delta  \mbox{ for $R \in \rel$}\\
\interp{\neg \varphi} & = & \Delta - \interp{\varphi}\\
\interp{\varphi \sqcap \varphi'} & = & \interp{\varphi} \cap \interp{\varphi'}\\
\interp{\exists R.\varphi} & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}\\
& & \mbox{ and } i' \in \interp{\varphi} \}.\\
\end{array}
$$

@@

Different description logics differ in the inventory of logical
connectives they allow: While \alc\ permits negation, \el\
doesn't. There are many other description logics in the literature;
some that we will get back to in Section~\ref{sec:related} are
$\mathcal{CL}$ (\el\ without existential quantification, i.e., only
conjunctions of atoms); $\mathcal{PL}$ (\alc\ without existential
quantification, i.e., propositional logic); and
$\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic negation).

\todo{formula preservation} A key notion in description logics that we
will use here is \emph{similarity}.  For any description logic $\gL$,
we will say that an individual $i$ is \emph{\gL-similar} to $i'$ in a
given model $\gM$ if for any formula $\varphi \in \gL$ such that $i
\in \interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
Equivalently, there is no $\gL$-formula that holds of $i$ but not of
$i'$.  We say that the \emph{\gL-similarity set} of some individual
$i$ is the set of all individuals to which $i$ is \gL-similar.

Notice that similarity is not necessarily a symmetrical relation: For
instance, $f_1$ is \el-similar to $f_2$ in
Fig.~\ref{fig:dale-haddock}b, but $f_2$ is not \el-similar to $f_1$
(it satisfies the formula $\exists \mathsf{in}.\mathsf{hat}$ and $f_1$
doesn't).  However, \alc-similarity is a symmetrical relation because
the language contains negation; and indeed, $f_1$ is not \alc-similar
to $f_2$ either because it satisfied $\neg \exists
\mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
it is generally possible for some individual $a$ to be \el-similar but
not \alc-similar to some individual $b$, but not vice versa.

@@

\begin{definition}
Given a model $\gM = (\Delta,\interp{\cdot})$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
\subseteq \Delta \times \Delta$ such that 
\begin{itemize}
\item[i)] $(i,i') \in Z$.\\[-1.5em]
\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
$(e_1,e_1') \in \interp{R}$ then there is $e_2'$ such that $(e_2,e_2') \in \interp{R}$ and
$(e_1',e_2') \in Z$.\\[-1.5em]
\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
$(e_2,e_2') \in \interp{R}$ then there is $e_1'$ such that $(e_1,e_1') \in \interp{R}$ and
$(e_1',e_2') \in Z$.
\end{itemize}
%A maximal subset of $\Delta$ all of whose elements are pairwise
%bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
\end{definition}

### 2008-03-21 trunk/Papers/dl-gre-08/conclusion.tex
edits on the phone

In this paper, we have explored the idea of viewing the generation of
singular REs as the problem of computing a DL formula with a given
extension.  We have shown how such formulas can be computed
efficiently (for \alc\ and \el) by adapting existing algorithms for
computing similarity sets.  The \el\ algorithm is able to generate
95\% of the non-relational and 67\% of the relational REs from
\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
are extremely efficient (350 ms and 140 ms respectively to generate
relational REs for all individuals in a random model with 100
individuals); to our knowledge, these are by far the fastest runtimes
for relational GRE reported in the literature.  We will make our
implementation available online.

Because they compute referring expressions for all individuals in the
domain at once, our algorithms will perform especially strongly in
static settings, such as the generation of descriptions for museum
exhibits, in which the individuals and their properties don't change
much.  However, even in more dynamic settings, our algorithms have a
chance to outperform search algorithms like Dale and Haddock's in the
average case because they can't get stuck in unproductive branches of
the search space. Nevertheless, one interesting question for future
research is how to incrementally update simulation classes when the
model changes. Similarly, it would be interesting to explore how
linguistic constraints and different attribute orderings can be taken
into account without slowing the algorithms down, and how the
algorithms can be adapted to compute REs for sets. In exploring these
extensions we will be able to draw on a rich body of literature that
considers many variants of simulation algorithms, including for
example incremental simulation algorithms or algorithms for
identifying sets of individuals.

>>>>

In this paper, we have explored the idea of viewing the generation of
singular REs as the problem of computing a DL formula with a given
extension.  We have shown how such formulas can be computed
efficiently (for \alc\ and \el) by adapting existing algorithms from
the literature.  The \el\ algorithm is able to generate 95\% of the
non-relational and 67\% of the relational REs from
\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
are extremely efficient (350 ms and 140 ms respectively to generate
relational REs for all individuals in a random model with 100
individuals); to our knowledge, these are by far the fastest runtimes
for relational GRE reported in the literature.  We will make our
implementation available online.

Because they compute referring expressions for all individuals in the
domain at once, our algorithms will perform especially strongly in
static settings, such as the generation of descriptions for museum
exhibits, in which the individuals and their properties don't change
much.  However, even in more dynamic settings, our algorithms have a
chance to outperform search algorithms like Dale and Haddock's in the
average case because they can't get stuck in unproductive branches of
the search space. Nevertheless, one interesting question for future
research is how to incrementally update simulation classes when the
model changes. Similarly, it would be interesting to explore how
linguistic constraints and different attribute orderings can be taken
into account without slowing the algorithms down, and how the
algorithms can be adapted to compute REs for sets. In exploring these
extensions we will be able to draw on a rich body of literature that
considers many variants of simulation algorithms.

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex
edits on the phone

To compare the descriptions generated by our algorithm to those humans
produce, we use a corpus of human-generated referring expressions
collected and made available by Jette Viethen and Robert
Dale.\footnote{http://www.ics.mq.edu.au/\~{}jviethen/drawers}  They
asked human subjects to describe one of 16 filing cabinet drawers. The
drawers had different colors and were arranged in a four-by-four grid
(see Fig.~\ref{fig:drawers}). The human subjects used seven
non-relational properties (the drawer's color, its $\mathsf{column}$
and $\mathsf{row}$ number, and whether it is in a $\mathsf{corner}$)
and five relational properties ($\mathsf{above, below, next\ to, left\
  of, right\ of}$). Of the 118 referring expressions, only 15 use
relations.

@@

We replicated Viethen and Dale's experiment for the \el\ algorithm
presented above. In the non-relational case, our results are the same
as theirs for the Incremental Algorithm: the \el\ algorithm generates
98 of the 103 non-relational description.  This is because the two
algorithms perform essentially the same computations if there are no
relations.

To evaluate the relational case, we first removed the
$\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
because these predicates make it possible to uniquely identify each
drawer with non-relational descriptions, and our algorithm would
always prefer these.  Using three (of 240) different orderings, our
algorithm was able to generate 10 of the 15 human-produced relational
descriptions correctly.  This is shown in more detail in
Fig.~\ref{fig:example_outputs}; notice that four of the 15
descriptions occurred twice in the corpus.  Of the five human-produced
descriptions that the \el\ algorithm cannot generate, two mention
column information, which we excluded from our domain, and three
involve references to sets (\textit{the two blues ones in horizontal
sequence}/\textit{the two yellow drawers}). While our algorithm cannot
reproduce those descriptions, it does generate other, simpler
descriptions for them.

>>>>

To compare the descriptions generated by our algorithm to those humans
produce, we use a corpus of human-generated referring expressions
collected and made available by Jette Viethen and Robert
Dale.\footnote{http://www.ics.mq.edu.au/\~{}jviethen/drawers} They
asked human subjects to describe one of 16 filing cabinet drawers. The
drawers had different colors and were arranged in a four-by-four grid
(see Fig.~\ref{fig:drawers}). The human subjects used four
non-relational properties (the drawer's $\mathsf{color}$, its
$\mathsf{column}$ and $\mathsf{row}$ number, and whether it is in a
$\mathsf{corner}$) and five relational properties ($\mathsf{above,
  below, next\ to, left\ of, right\ of}$). Of the 118 referring
expressions obtained in the experiment, only 15 use relations.

@@

We replicated Viethen and Dale's experiment for the \el\ algorithm
presented above. In the non-relational case, our results are the same
as theirs for the Incremental Algorithm: the \el\ algorithm generates
98 of the 103 non-relational descriptions, using four (of the
possible) orderings.  This is because the two algorithms perform
essentially the same computations if there are no relations.

To evaluate the relational case, we first removed the
$\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
because these predicates make it possible to uniquely identify each
drawer with non-relational descriptions, and our algorithm would
always prefer these.  Using three (of the 240 possible) different
orderings, our algorithm was able to generate 10 of the 15
human-produced relational descriptions correctly.  This is shown in
more detail in Fig.~\ref{fig:example_outputs}; notice that four of the
15 descriptions occurred twice in the corpus.  Of the five
human-produced descriptions that the \el\ algorithm cannot generate,
two mention column information, which we excluded from our domain, and
three involve references to sets (\textit{the two blues ones in
  horizontal sequence}/\textit{the two yellow drawers}). While our
algorithm cannot reproduce those descriptions, it does generate other,
simpler descriptions for them.

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
edits on the phone

\medskip
\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that
$\interp{\varphi}^\gM = A$\\ 
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

In the examples above, it is because $\mathsf{flower} \sqcap \exists
\mathsf{in}. \mathsf{hat}$ denotes exactly $\{f_2\}$ that we can say
``the flower in the hat'' to refer to $f_2$.  This perspective
provides a general framework into which many existing GRE approaches
fit: Traditional attribute selection \cite{Dale1995} corresponds to
building DL formulas that are conjunctions of atoms; relational REs as
in \newcite{dale91:_gener_refer_expres_invol_relat} are formulas of
\el; and so on.  We will pursue the idea of organizing GRE approaches
with respect to the variant of DL they use further in
Section~\ref{sec:related}.

@@

In the rest of this section, we will present algorithms that compute
the similarity sets of a given model for \alc\ and \el, and for each
similarity set, a characteristic formula that denotes exactly this
set.  In the \alc\ case, we adapt a standard algorithm from the
literature for computing \emph{simulation classes}; we will then
further adapt this algorithm for \el.  In effect, both algorithms
compute REs for all individuals in some model at the same time -- very
efficiently and without any danger of infinite regress.

@@

It can be shown that for \alc, the similarity sets of a finite model
coincide exactly with the \emph{simulation classes} of this model
\cite{blac:moda01}.  We don't have the space here to define simulation
classes, but they have been studied extensively in the literature, and
there are several efficient algorithms for computing \alc-simulation
classes
\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
However, these algorithms will only compute the simulation classes
themselves. Here we extend the \newcite{hopc:algo71} algorithm such
that it computes, along with each set, also a formula that denotes
exactly this set.  We can then use these formulas as representations
of the REs.

@@

The \alc\ algorithm computes the \alc-similarity sets of the model in
time $O(nm)$, where $n$ is the number of individuals in the domain,
and $m$ is the number of pairs of individuals that are related by some
relation.  However, it will freely introduce negations in the case
distinctions, which can make the resulting formula hard to realize
(see also Section~\ref{sec:discussion-realization}).  This is why we
also present an algorithm for the \el-similarity sets; \el\
corresponds to positive relational REs, which are generally much
easier to realize.

We obtain the \el\ algorithm by replacing the call to add$_{\alc}$ in
Algorithm~\ref{algo:bisim-l} by a call to add$_{\el}$, which is
defined in Algorithm~\ref{algo:bisim-add-el}.  As before, the
algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
formulas (this time of \el) such that $\interp{\varphi_1} \cup \ldots
\cup \interp{\varphi_n} = \Delta$, and which it refines iteratively.
However, where the \alc\ algorithm maintains the invariant that
$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of
$\Delta$, we weaken this invariant to the requirement that there are
no $m \geq 2$ pairwise different indices $1 \leq i_1,\ldots,i_m \leq
n$ such that $\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup
\ldots \cup \interp{\varphi_{i_m}}$.  We call the formula
$\varphi_{i_1}$ \emph{subsumed} if such a decomposition exists.

@@

The \el\ algorithm has a harder time with the example in
Fig.~\ref{fig:dale-haddock}b \cite{Stone1998a}.  While it will
correctly identify $r_1$ as ``the rabbit in the hat'' and $f_2$ as
``the flower in the bathtub'', it will not be able to compute a RE for
$f_1$ because $f_1$ is \el-similar to $f_2$.  Indeed, the algorithm
terminates with $\RE$ containing both $\mathsf{flower}$ and
$\mathsf{flower} \sqcap \exists \mathsf{in}.\mathsf{hat}$.  This is a
typical pattern for asymmetrical cases of similarity in \el: If there
are two formulas $\varphi_1$ and $\varphi_2$ in the output set with
$\interp{\varphi_1} \subseteq \interp{\varphi_2}$, then there is
generally some individual $b \in \interp{\varphi_2} -
\interp{\varphi_1}$ such that all individuals in $\interp{\varphi_1}$
are similar to $b$, but not vice versa.  By contrast, the \alc\
algorithm can exploit the greater expressivity of \alc\ to split
$\mathsf{flower}$ into the two new formulas $\mathsf{flower} \sqcap
\exists \mathsf{in}.\mathsf{hat}$ and $\mathsf{flower} \sqcap \neg
\exists \mathsf{in}.\mathsf{hat}$, generating a unique RE for $f_1$ as
well.

>>>>

\medskip
\noindent
{\small
\begin{center}
\begin{tabular}{ll} \hline
\multicolumn{2}{l}{
\textsc{$\gL$-GRE Problem}}\\ \hline
\ \ Input: & A model $\gM$ and a target set $A \subseteq \Delta$.\\
\ \ Output: & A formula $\varphi \in \gL$ such that
$\interp{\varphi} = A$\\ 
& \hspace*{0.5cm} (if such a formula exists).\\ \hline
\end{tabular}
\end{center}}
\medskip

In the examples above, it is because $\mathsf{flower} \sqcap \exists
\mathsf{in}. \mathsf{hat}$ denotes exactly $\{f_2\}$ that we can say
``the flower in the hat'' to refer to $f_2$.  This perspective
provides a general framework into which many existing GRE approaches
fit: Traditional attribute selection \cite{Dale1995} corresponds to
building DL formulas that are conjunctions of atoms; relational REs as
in \newcite{dale91:_gener_refer_expres_invol_relat} are formulas of
\el; and so on.  We will further pursue the idea of organizing GRE
approaches with respect to the variant of DL they use in
Section~\ref{sec:related}.

@@

In the rest of this section, we will present algorithms that compute
the similarity sets of a given model for \alc\ and \el, together with
characteristic formulas that denote them.  In the \alc\ case, we adapt
a standard algorithm from the literature for computing
\emph{simulation classes}; we will then further adapt this algorithm
for \el.  In effect, both algorithms compute REs for all individuals
in some model at the same time -- very efficiently and without any
danger of infinite regress.

@@

It can be shown that for \alc, the similarity sets of a finite model
coincide exactly with the \emph{simulation classes} of this model.
Simulation classes have been studied extensively in the literature
\cite{blac:moda01,kurt:expr99}, and there are several efficient
algorithms for computing \alc-simulation classes
\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
However, these algorithms will only compute the simulation classes
themselves. Here we extend the \newcite{hopc:algo71} algorithm such
that it computes, along with each set, also a formula that denotes
exactly this set.  We can then use these formulas as representations
of the REs.

@@

The \alc\ algorithm computes the \alc-similarity sets of the model in
time $O(n^3)$, where $n$ is the number of individuals in the domain.
However, it will freely introduce negations in the case distinctions,
which can make the resulting formula hard to realize (see also
Section~\ref{sec:discussion-realization}).  This is why we also
present an algorithm for the \el-similarity sets; \el\ corresponds to
positive relational REs, which are generally much easier to realize.

We obtain the \el\ algorithm by replacing the call to add$_{\alc}$ in
Algorithm~\ref{algo:bisim-l} by a call to add$_{\el}$, which is
defined in Algorithm~\ref{algo:bisim-add-el}.  As before, the
algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
formulas (this time of \el) such that $\interp{\varphi_1} \cup \ldots
\cup \interp{\varphi_n} = \Delta$, and which it refines iteratively.
However, where the \alc\ algorithm maintains the invariant that
$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of
$\Delta$, we weaken this invariant to the requirement that there are
no $m \geq 2$ pairwise different indices $1 \leq i_1,\ldots,i_m \leq
n$ such that $\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup
\ldots \cup \interp{\varphi_{i_m}}$.  We call $\varphi_{i_1}$
\emph{subsumed} if such a decomposition exists.

@@

The \el\ algorithm has a harder time with the example in
Fig.~\ref{fig:dale-haddock}b \cite{Stone1998a}.  While it will
correctly identify $r_1$ as ``the rabbit in the hat'' and $f_2$ as
``the flower in the hat'', it will not be able to compute a RE for $f_1$
because $f_1$ is \el-similar to $f_2$.  Indeed, the algorithm
terminates with $\RE$ containing both $\mathsf{flower}$ and
$\mathsf{flower} \sqcap \exists \mathsf{in}.\mathsf{hat}$.  This is a
typical pattern for asymmetrical cases of similarity in \el: If there
are two formulas $\varphi_1$ and $\varphi_2$ in the output set with
$\interp{\varphi_1} \subseteq \interp{\varphi_2}$, then there is
generally some individual $b \in \interp{\varphi_2} -
\interp{\varphi_1}$ such that all individuals in $\interp{\varphi_1}$
are similar to $b$, but not vice versa.  By contrast, the \alc\
algorithm can exploit the greater expressivity of \alc\ to split
$\mathsf{flower}$ into the two new formulas $\mathsf{flower} \sqcap
\exists \mathsf{in}.\mathsf{hat}$ and $\mathsf{flower} \sqcap \neg
\exists \mathsf{in}.\mathsf{hat}$, generating a unique RE for $f_1$ as
well.

### 2008-03-21 trunk/Papers/dl-gre-08/introduction.tex
edits on the phone

Referring expressions involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
and Haddock algorithm and most of its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
vulnerable to the problem of ``infinite regress'', where the algorithm
jumps back and forth between generating descriptions for two related
individuals infinitely, as in ``the book on the table which supports a
book on the table \ldots''.

In this paper, we propose to view GRE as the problem of computing a
formula of description logic (DL) that denotes exactly the set of
individuals that we want to refer to.  This very natural idea has been
mentioned in passing before
\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
take it one step further by proposing DL as an interlingua for
comparing the referring expressions (REs) produced by different
approaches to GRE.  In this way, we can organize existing GRE
approaches in an expressiveness hierarchy.  For instance, the
classical Dale and Reiter algorithms compute purely conjunctive
formulas;
\newcite{deemter01:_gener_refer_expres} extends this language by
adding the other propositional connectives, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
allowing existential quantification.

>>>>

Referring expressions (REs) involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
and Haddock algorithm and most of its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
vulnerable to the problem of ``infinite regress'', where the algorithm
jumps back and forth between generating descriptions for two related
individuals infinitely, as in ``the book on the table which supports a
book on the table \ldots''.

In this paper, we propose to view GRE as the problem of computing a
formula of description logic (DL) that denotes exactly the set of
individuals that we want to refer to.  This very natural idea has been
mentioned in passing before
\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
take it one step further by proposing DL as an interlingua for
comparing the REs produced by different approaches to GRE.  In this
way, we can organize existing GRE approaches in an expressiveness
hierarchy.  For instance, the classical Dale and Reiter algorithms
compute purely conjunctive formulas;
\newcite{deemter01:_gener_refer_expres} extends this language by
adding the other propositional connectives, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
allowing existential quantification.

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex
edits on the phone

Viewing GRE as a problem of generating DL formulas offers a new,
unified perspective of the GRE problem: It is the problem of computing
a DL formula with a certain (typically singleton) extension.  Many
existing approaches can be subsumed under this view; we have
summarized this for some of them in Fig.~\ref{fig:related}, along with
the DL fragment they use.  We already discussed some of these
approaches in Section~\ref{sec:gre}.  Furthermore, the non-relational
but negative and disjunctive descriptions generated by
\newcite{deemter01:_gener_refer_expres} are simply formulas of
$\mathcal{PL}$; and \newcite{gardent02:_gener_minim_defin_descr}
generalizes this into generating formulas of $\mathcal{ELU}_{(\neg)}$,
i.e., \el\ plus disjunction and atomic negation.
The approach presented here fits well into this landscape, and it
completes the picture by showing how to generate formulas for
arbitrary bisimulation classes in \alc, which combines all connectives
used in any of these previous approaches.

>>>>

Viewing GRE as a problem of generating DL formulas offers a unified
perspective of the GRE problem: It is the problem of computing a DL
formula with a given extension.  Many existing approaches can be
subsumed under this view; we have summarized this for some of them in
Fig.~\ref{fig:related}, along with the DL fragment they use.  We
already discussed some of these approaches in Section~\ref{sec:gre}.
Furthermore, the non-relational but negative and disjunctive
descriptions generated by \newcite{deemter01:_gener_refer_expres} are
simply formulas of $\mathcal{PL}$; and
\newcite{gardent02:_gener_minim_defin_descr} generalizes this into
generating formulas of $\mathcal{ELU}_{(\neg)}$, i.e., \el\ plus
disjunction and atomic negation.  The approach presented here fits
well into this landscape, and it completes the picture by showing how
to generate REs in \alc, which combines all connectives used in any of
these previous approaches.
